---
title: "Characterising CLL bioenergetics through Seahorse extracellular flux assay"
vignette: >
  %\VignetteIndexEntry{seahorseCLL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---


# Introduction

This package contains the data of bioenergetic features of 140 CLL samples and 9 B cell samples, measured by Seahorse extracellular flux assay. This package also reproduces figures presented in the paper "Energy metabolism is co-determined by genetic variants in chronic lymphocytic leukemia and influences drug sensitivity" by Lu J, Böttcher M et al.

***

In this vignette we present the integrative analysis of CLL bioenergetic data set and the source code for the paper.  

This vignette is build from the sub-vignettes, which each can be build separatelly. The parts are separated by the horizontal lines. Each part finishes with removal of all the created objects.


```{r echo=FALSE}
.standalone=FALSE
```

```{r, message=FALSE, warning=FALSE, include=!.standalone}
library(seahorseCLL)
library(BloodCancerMultiOmics2017)
library(SummarizedExperiment)
library(ggbeeswarm)
library(xtable)
library(cowplot)
library(piano)
library(gridExtra)
library(grid)
library(genefilter)
library(pheatmap)
library(ggrepel)
library(GEOquery)
library(limma)
library(robustbase)
library(DESeq2)
library(survival)
library(maxstat)
library(survminer)
library(glmnet)
library(RColorBrewer)
library(reshape2)
library(gtable)
library(Biobase)
library(tidyverse)
```

***

<!-- **p1** -->

```{r p01, child = 'src/section01.Rmd'}
```

***

<!-- **p2** -->

```{r p02, child = 'src/section02.Rmd'}
```

***

<!-- **p3** -->

```{r p03, child = 'src/section03.Rmd'}
```

***

<!-- **p4**-->

```{r p04, child = 'src/section04.Rmd'}
```

***

<!-- **p5** -->

```{r p05, child = 'src/section05.Rmd'}
```

***

<!-- **p6** -->

```{r p06, child = 'src/section06.Rmd'}
```

***

# End of session

```{r}
sessionInfo()
```
