---
title: 'Section 5: Association between the clinical phenotype and energy metabolic features'
author: "Junyan Lu"
date: "`r doc_date()`"
output:
  BiocStyle::html_document
---

# Association between the clinical phenotype and energy metabolic features

```{r, message=FALSE, include=!exists(".standalone"), eval=!exists(".standalone"), echo = FALSE}
library(seahorseCLL)
library(BloodCancerMultiOmics2017)
library(SummarizedExperiment)
library(grid)
library(survival)
library(gridExtra)
library(maxstat)
library(xtable)
library(cowplot)
library(ggbeeswarm)
library(survminer)
library(tidyverse)
```

```{r echo=FALSE}
plotDir = ifelse(exists(".standalone"), "", "section05/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
options(stringsAsFactors=FALSE)
```



## Correlation between seahorse measurement and pretreatment status

### Correlation test

Load data set
```{r}
data("patmeta", "seaCombat","lpdAll", "pretreat","doublingTime")
```

Prepare data table for t-test
```{r}
testTab <- assays(seaCombat)$seaMedian %>% data.frame() %>% rownames_to_column("Measurement") %>%
  gather(key = "patID", value = "value", -Measurement) %>%
  mutate(pretreated = pretreat[patID,]) %>%
  mutate(pretreated = factor(pretreated),
         IGHV = factor(Biobase::exprs(lpdAll)["IGHV Uppsala U/M",patID])) %>% 
  as.tibble()
```

Performing t-test for each measurement
```{r}
tTest <- function(x, y) {
  noNA <- !is.na(x) & !is.na(y)
  res <- t.test(x[noNA] ~ y[noNA], equal.var = TRUE)
  tibble(p = res$p.value,
         dm = res$estimate[[2]] - res$estimate[[1]])
}

pTab <- group_by(testTab, Measurement) %>% do(tTest(.$value, .$pretreated)) %>% ungroup() %>%
  mutate(p.adj = p.adjust(p, method = "BH"))
```


Perform ANOVA test, including IGHV as a blocking factor
```{r}
aovTest <- function(x,y,block) {
  noNA <- !is.na(x) & !is.na(y) & !is.na(block)
  dataTab <- data.frame(x = x[noNA], y = y[noNA], block = block[noNA])
  res <- anova(lm(x ~ y + block))
  tibble(p = res$`Pr(>F)`[1])
}

pTab.IGHV <- group_by(testTab, Measurement) %>% do(aovTest(.$value, .$pretreated, .$IGHV)) %>%
  ungroup() %>% mutate(p.adj = p.adjust(p, method = "BH"))
```

A table as output
```{r, results=FALSE}
outTable <- tibble(`Seahorse mearuement` = formatSea(pTab$Measurement),
                   `p value` = format(pTab$p, digits = 2),
                   `adjusted p` = format(pTab$p.adj, digits = 3),
                   `p value (IGHV blocked)` = format(pTab.IGHV$p, digits = 2),
                   `adjusted p (IGHV blocked)` = format(pTab.IGHV$p.adj, digits = 3))

write(
print(xtable(outTable, digits = 3, 
             caption = "Student's t-test and ANOVA test (IGHV blocked) results of energy metabolic measurements related to pretreatment status"), 
      include.rownames=FALSE,
      caption.placement = "top")
,file = "section05/tTest_SeahorseVSpretreat.tex")
```

Association between IGHV status and pretreatment
```{r}
IGHVtab <- filter(testTab, !duplicated(patID))
chiRes <- chisq.test(IGHVtab$pretreated, IGHVtab$IGHV)
chiRes
```



### Plot significant associations as beeswarms plot (for supplementary figures)

Plot
```{r}
seaList <- c("maximal.respiration", "spare.respiratory.capacity")
plotList <- lapply(seaList, function(seaName) {
  #prepare table for plotting
  plotTab <- filter(testTab, Measurement == seaName) %>% 
    mutate(Measurement = formatSea(Measurement),
           pretreated = ifelse(pretreated ==1, 
                               sprintf("yes (n=%s)", sum(pretreated == 1)), 
                               sprintf("no (n=%s)", sum(pretreated == 0))),
           IGHV = ifelse(is.na(IGHV), "unknown", 
                         ifelse(IGHV == 1, "mutated", "unmutated"))) %>%
    mutate(IGHV = factor(IGHV, levels = c("mutated","unmutated","unknown"))) %>%
    filter(!is.na(value))
  
  #p value for annotation
  pval <- filter(pTab, Measurement == seaName)$p
  
  #color for IGHV status
  colorList <- c(mutated = "red", unmutated = "black", unknown = "grey80")
  
  #formatted seaname
  seaTitle <- unique(plotTab$Measurement)
  
  #title 
  plotTitle <- paste(sprintf("'%s (p = '~",seaTitle),
                     sciPretty(pval, digits = 2),"*')'")
  
  ggplot(plotTab, aes(x=pretreated, y = value)) + 
    stat_boxplot(geom = "errorbar", width = 0.3) +
    geom_boxplot(outlier.shape = NA, col="black", width=0.4) + 
    geom_beeswarm(cex=2, size =1, aes(col = IGHV)) + theme_classic() +
    xlab("") + ylab("OCR (pMol/min)") + ggtitle(parse(text=plotTitle)) + 
    scale_color_manual(values = colorList) +
    theme(axis.line.x = element_blank(), axis.ticks.x = element_blank(),
           axis.title = element_text(size=12, face="bold"),
           axis.text = element_text(size=12),
           plot.title = element_text(hjust=0.5,size=15),
           axis.title.x = element_text(face="bold"))
    
})
```

Save the plot
```{r Seahorse_vs_pretreat, fig.height=5, fig.width=12}
grid.arrange(grobs = plotList, ncol =2)
```

## Correlation between seahorse measurement and lymphocyte doubling time

### Correlation test

Pre-processing data
```{r}
testTab <- assays(seaCombat)$seaMedian %>% data.frame() %>% rownames_to_column("Measurement") %>%
  gather(key = "patID", value = "value", -Measurement) %>%
  mutate(pretreated = pretreat[patID,],
         doubling.time = LDT[patID,]) %>%
  mutate(pretreated = factor(pretreated),
         IGHV = factor(Biobase::exprs(lpdAll)["IGHV Uppsala U/M",patID])) %>% 
  filter(!is.na(doubling.time), !is.na(value)) %>%
  as.tibble()
```

Correlation test between seahorse measurement and doubling time
```{r}
corTest <- function(x, y, block = NULL, method = "pearson") {
  if (is.null(block)) {
    res <- cor.test(x,y, method = method)
    tibble(p = res$p.value, coef = res$estimate[[1]])
  } else {
    tab <- data.frame(x = x, y = y, block = block)
    res <- summary(lm( y ~ x + block, tab)) #how much y can be explained by x and block
    tibble(p = res$coefficients[2,4],
           coef = cor(x,y))
  }
}

corRes <- group_by(testTab, Measurement) %>% do(corTest(.$value, .$doubling.time)) %>%
  ungroup() %>% mutate(p.adj = p.adjust(p, method = "BH"))
```

Correlations between lymphocyte doubling time and IGHV stauts
```{r}
pRes <- filter(testTab, !duplicated(patID)) %>% 
  do(data.frame(p = t.test(doubling.time ~ IGHV,.)$p.value))
pRes
```

Correlation test between seahorse measurement and doubling time
(Considering IGHV as cofactor)
```{r}
corRes.aov <- group_by(testTab, Measurement) %>% filter(!is.na(IGHV)) %>%
  do(corTest(x = .$value, y = .$doubling.time, block = .$IGHV)) %>%
  ungroup() %>% mutate(p.adj = p.adjust(p, method = "BH"))
```

Correlation test between seahorse measurement and doubling time
(within M-CLL)
```{r}
corRes.M <- group_by(testTab, Measurement) %>% filter(IGHV %in% 1) %>%
  do(corTest(x = .$value, y = .$doubling.time)) %>%
  ungroup() %>% mutate(p.adj = p.adjust(p, method = "BH"))
```

Correlation test between seahorse measurement and doubling time
(within U-CLL)
```{r}
corRes.U <- group_by(testTab, Measurement) %>% filter(IGHV %in% 0) %>%
  do(corTest(x = .$value, y = .$doubling.time)) %>%
  ungroup() %>% mutate(p.adj = p.adjust(p, method = "BH"))
```

A table for output
```{r, results=FALSE}
outTable <- tibble(`Seahorse mearuement` = formatSea(corRes$Measurement),
                   `p value` = format(corRes$p, digits = 2),
                   `adjusted p` = format(corRes$p.adj, digits = 3),
                   `p value (IGHV blocked)` = format(corRes.aov$p, digits = 3),
                   `adjusted p (IGHV blocked)` = format(corRes.aov$p.adj, digits = 3))

write(
print(xtable(outTable, digits = 3, 
             caption = "Correlation tests between each Seahorse measurements and lymphocyte doubling time"), 
      include.rownames=FALSE,
      caption.placement = "top")
,"section05/tTest_SeahorseVSldt.tex")
```

### Plot correlations (for supplementary figure)

```{r}
seaList <- c("glycolysis", "glycolytic.capacity")
plotList.cor <- lapply(seaList, function(seaName) {
  #prepare table for plotting
  plotTab <- filter(testTab, Measurement == seaName) %>% 
    mutate(Measurement = formatSea(Measurement),
           IGHV = ifelse(is.na(IGHV), "unknown", 
                         ifelse(IGHV == 1, "mutated", "unmutated"))) %>%
    mutate(IGHV = factor(IGHV, levels = c("mutated","unmutated","unknown"))) %>%
    filter(!is.na(value), !is.na(doubling.time))
  
  #p value for annotation
  pval <- filter(corRes, Measurement == seaName)$p
  coef <- filter(corRes, Measurement == seaName)$coef
  
  #color for IGHV status
  colorList <- c(mutated = "red", unmutated = "black", unknown = "grey80")
  
  #formatted seaname
  seaTitle <- unique(plotTab$Measurement)
  
  
  #prepare correlation test annotations
  annoText <- paste("'coefficient ='~",format(coef,digits = 2),"*","', p ='~",sciPretty(pval,digits=2))
  
  ggplot(plotTab, aes(x=value,y=doubling.time)) + 
    geom_point(size=1, aes(col = IGHV)) + 
    geom_smooth(method = "lm", se= FALSE) +
    xlab("ECAR (pMol/min)") + ylab("Lymphocyte doubling time (days)") +
    theme_bw() + ggtitle(seaTitle) +
    scale_color_manual(values = colorList) +
    annotate("text", x = -Inf, y = Inf, label = annoText, 
             vjust=1, hjust=0, size = 5, parse = TRUE, col= "darkred") +
    theme(panel.grid = element_blank(),
          axis.title = element_text(size=13, face="bold"),
          axis.text = element_text(size=12),
          plot.title = element_text(face="bold", hjust=0.5, size = 15 ),
          legend.position = "none",
          axis.title.x = element_text(face="bold"),
          plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
    
})
```

Show the plot
```{r Seahorse_vs_LDT, fig.height=5, fig.width=12}
grid.arrange(grobs = plotList.cor, ncol =2)
```

Combine and save the plot
```{r seahorse_clinics, fig.width=11, fig.height=10, fig.path=plotDir, dev=c("png", "pdf")}
plot_grid(plotList.cor[[1]], plotList.cor[[2]],
          plotList[[1]], plotList[[2]],labels = "AUTO")
```

## Correlation between seahorse measurement and survival

### Prepare dataset
```{r}
seaTable <- assays(seaCombat)$seaMedian

survT = patmeta[colnames(seaTable),]
survT[which(survT[,"IGHV"]=="U") ,"IGHV"] = 0
survT[which(survT[,"IGHV"]=="M") ,"IGHV"] = 1
survT$IGHV = as.numeric(survT$IGHV)

colnames(survT) = gsub("Age4Main", "age", colnames(survT))

#add seahorse measurement information
survT <- cbind(survT, t(seaTable))

# competinting risk endpoint fpr 
survT$compE <- ifelse(survT$treatedAfter == TRUE, 1, 0)
survT$compE <- ifelse(survT$treatedAfter == FALSE & survT$died==TRUE,
                      2, survT$compE )
survT$T7  <- ifelse(survT$compE == 1, survT$T5, survT$T6 )
```

### Univariate survival analysis

#### Caclulation correlations

Function to calculate correlations
```{r forest-together}

com <- function( Time, endpoint, scaleX, sub, d, split, drug_names) {  
  
  res <- lapply(d, function(g)  { 
  
  #drug <- survT[,g] * scaleX
  drug <- survT[,g]
  #drug <- (drug - mean(drug, na.rm = TRUE))/sd(drug, na.rm=TRUE)
  ## all=99, M-CLL=1, U-CLL=0
  if(sub==99) { surv <- coxph(Surv(survT[,paste0(Time)],
                                   survT[,paste0(endpoint)] == TRUE) ~ drug)} 
  if(sub<99)  { surv <- coxph(Surv(survT[,paste0(Time)],
                                   survT[,paste0(endpoint)] == TRUE) ~ drug,
                              subset=survT[,paste0(split)]==sub)}    
  
  c(summary(surv)[[7]][,5], summary(surv)[[7]][,2], 
                summary(surv)[[8]][,3], 
                summary(surv)[[8]][,4])
 })
 s <- do.call(rbind, res)
 colnames(s) <- c("p", "HR", "lower", "higher")
 rownames(s) <- drug_names
 s
}
```

All samples
```{r}
d <- rownames(seaTable)
sea_names <- formatSea(d)

ttt <- com(Time="T5", endpoint="treatedAfter", sub=99, d=d,
          split="IGHV", drug_names=sea_names, scaleX=1)

   
os <-  com(Time="T6", endpoint="died", sub=99, d=d, split="",
          drug_names=sea_names, scaleX=1)

#TTT result
ttt

#OS result
os
```

M-CLL samples
```{r}
d <- rownames(seaTable)
sea_names <- formatSea(d)

ttt.M <- com(Time="T5", endpoint="treatedAfter", sub=1, d=d,
          split="IGHV", drug_names=sea_names, scaleX=1)

   
os.M <-  com(Time="T6", endpoint="died", sub=1, d=d, 
           split="IGHV",drug_names=sea_names, scaleX=1)

#TTT result
ttt.M

#OS result
os.M
```


U-CLL samples
```{r}
d <- rownames(seaTable)
sea_names <- formatSea(d)

ttt.U <- com(Time="T5", endpoint="treatedAfter", sub=0, d=d,
          split="IGHV", drug_names=sea_names, scaleX=1)

   
os.U <-  com(Time="T6", endpoint="died", sub=0, d=d, 
           split="IGHV",drug_names=sea_names, scaleX=1)

#TTT result
ttt.U

#OS result
os.U
```

#### KM plot

Function for km plot
```{r}
#need to be fixed
km <- function(survT, seaName, split, titlePlot, t, hr, c, pvals = NULL) { 
  #function for km plot
  survS <- survT
    #filter_(survT, sprintf("!is.na(%s)",split),
    #              sprintf("!is.na(%s)",seaName))
  k <- survS[ , seaName]
  
  ms5 <- maxstat.test(Surv(T5, treatedAfter)  ~ k, 
                             data = survS,
                             smethod = "LogRank",
                             minprop = 0.2, 
                             maxprop = 0.8, 
                             alpha = NULL)
  ms6 <- maxstat.test(Surv(T6, died) ~ k, 
                             data = survS,
                             smethod = "LogRank",
                             minprop = 0.2, 
                             maxprop = 0.8, 
                             alpha = NULL)

  med <- median(k, na.rm = TRUE)
  
  # median
  if (c=="med") {    
   survS$cutA <- ifelse(k >= med, "high", "low")
   survS$cutM <- ifelse(k >= med, "high", "low")
   survS$cutU <- ifelse(k >= med, "high", "low")
  }
  
  #maxstat & TTT
  if (c=="maxstat" & t=="TTT") {    
   survS$cutA <- ifelse(k >= ms5$estimate, "high", "low")
   survS$cutM <- ifelse(k >= ms5$estimate, "high", "low") 
   survS$cutU <- ifelse(k >= ms5$estimate, "high", "low")
  }
  
  #OS & maxstat
  if (c=="maxstat" & t=="OS") {    
   survS$cutA <- ifelse(k >= ms6$estimate, "high", "low")
   survS$cutM <- ifelse(k >= ms6$estimate, "high", "low") 
   survS$cutU <- ifelse(k >= ms6$estimate, "high", "low")
  }
  
  p <- list()  #list for storing plots
  
  #prepare p value annotations
  if (is.null(pvals)){
    pvalList <- rep(FALSE,3)
  } else {
    pvalList <- pvals
  }
  
  #subset according to genetic factor
  survM <- survS[survS[,split] %in% 1,]
  survU <- survS[survS[,split] %in% 0,]
  inputName <- formatSea(seaName)
  if (t=="TTT") {
         yl <- "Fraction w/o treatment"
         #if (c=="med"){ cat(sprintf("%s median-cutpoint for TTT: %5.2f\n", inputName, median(k) ) ) } else 
         #  { cat(sprintf("%s cutpoint for TTT: %5.2g\n", inputName, ms5$estimate )) }
         p[[1]] <- ggsurvplot(survfit(Surv(T5, treatedAfter) ~ cutA, data = survS), 
                              data = survS, pval = pvalList[1],
                              ylab = yl, xlab = "Time (years)", title = inputName,
                              legend.labs = c("high","low"), palette = c("red","blue"),
                              ggtheme = theme(plot.title = element_text(vjust =0.5, size = 18),
                                              axis.text = element_text(size= 15)))$plot
         p[[2]] <- ggsurvplot(survfit(Surv(T5, treatedAfter) ~ cutM, data = survM),
                              data = survS, pval = pvalList[2],
                              ylab = yl, xlab = "Time (years)",
                              title = paste(inputName, titlePlot[1], titlePlot[3]),
                              legend.labs = c("high","low"), palette = c("red","blue"),
                              ggtheme = theme(plot.title = element_text(vjust =0.5, size = 18),
                                              axis.text = element_text(size= 15)))$plot
         p[[3]] <- ggsurvplot(survfit(Surv(T5, treatedAfter) ~ cutU, data = survU),
                              data = survS, pval = pvalList[3],
                              ylab = yl, xlab = "Time (years)", 
                              title = paste(inputName, titlePlot[1], titlePlot[2]),
                              legend.labs = c("high","low"), palette = c("red","blue"),
                              ggtheme = theme(plot.title = element_text(vjust =0.5, size = 18),
                                              axis.text = element_text(size= 15)))$plot
         }
   # OS  
   else {
         yl <- "Fraction overall survival"
         #if (c=="med"){ cat(sprintf("%s median-cutpoint for OS: %5.2f\n", inputName, median(k) ) ) } else 
         #  { cat(sprintf("%s cutpoint for OS: %5.2f\n", inputName, ms6$estimate )) }
         p[[1]] <- ggsurvplot(survfit(Surv(T5, died) ~ cutA, data = survS), data = survS,
                              ylab = yl, xlab = "Time (years)", pval = pvalList[1],
                              title = inputName, legend.labs = c("high","low"), palette = c("red","blue"),
                              ggtheme = theme(plot.title = element_text(vjust =0.5, size = 18),
                                              axis.text = element_text(size= 15)))$plot
         p[[2]] <- ggsurvplot(survfit(Surv(T5, died) ~ cutM, data = survM), data = survM,
                              ylab = yl, xlab = "Time (years)", pval = pvalList[2],
                              title = paste(inputName, titlePlot[1], titlePlot[3]), 
                              legend.labs = c("high","low"), palette = c("red","blue"),
                              ggtheme = theme(plot.title = element_text(vjust =0.5, size = 18),
                                              axis.text = element_text(size= 15)))$plot 
         p[[3]] <- ggsurvplot(survfit(Surv(T5, died) ~ cutU, data = survU), data = survU,
                              ylab = yl, xlab = "Time (years)", pval = pvalList[3],
                              title = paste(inputName, titlePlot[1], titlePlot[2]), 
                              legend.labs = c("high","low"), palette = c("red","blue"),
                              ggtheme = theme(plot.title = element_text(vjust =0.5, size = 18),
                                              axis.text = element_text(size= 15)))$plot 
   }
  p
}
```

#### Using maxstat

Time to next treatment (maxstat).
```{r}
TTTplot.maxstat <- list()

for (seaName in rownames(seaCombat)) {
  seaName.format <- formatSea(seaName)
  pvals <- c(ttt[seaName.format ,"p"], ttt.M[seaName.format ,"p"], ttt.U[seaName.format ,"p"])
  pvals <- sapply(pvals, function(x) {
    sprintf("p value = %1.2f",x)
  })
  TTTplot.maxstat[[seaName]] <- km(survT, seaName = seaName, split = "IGHV", t="TTT",
   titlePlot =c("(IGHV", "umutated)", "mutated)"),  hr="tr", c="maxstat", pvals = pvals)
}
```

Overall survival (maxstat).
```{r, fig.path=plotDir, fig.width = 10, fig.height = 3.3, dev = c("png", "pdf")}
OSplot.maxstat <- list()

for (seaName in rownames(seaCombat)) {
  seaName.format <- formatSea(seaName)
  pvals <- c(os[seaName.format ,"p"], os.M[seaName.format ,"p"], os.U[seaName.format ,"p"])
  pvals <- sapply(pvals, function(x) {
    sprintf("p value = %1.2f",x)
  })
  OSplot.maxstat[[seaName]] <- km(survT, seaName = seaName, split = "IGHV", t="OS",
   titlePlot =c("(IGHV", "umutated)", "mutated)"),  hr="tr", c="maxstat", pvals = pvals)
}
```

KM plot for all samples (not shown in manuscript)
```{r TTT_OS_combine, fig.path=plotDir, fig.width = 12, fig.height = 8, dev = c("png", "pdf")}
plot_grid(TTTplot.maxstat$glycolytic.reserve[[1]], TTTplot.maxstat$maximal.respiration[[1]],
          TTTplot.maxstat$spare.respiratory.capacity[[1]],
          OSplot.maxstat$glycolytic.capacity[[1]], OSplot.maxstat$glycolytic.reserve[[1]],
          labels = "AUTO")
```

KM plot for IGHV stratified samples (not shown in manuscript)
```{r TTT_OS_IGHV_UM, fig.path=plotDir, fig.width = 10, fig.height = 20, dev = c("png", "pdf")}
plot_grid(TTTplot.maxstat$glycolytic.reserve[[2]], TTTplot.maxstat$glycolytic.reserve[[3]],
          TTTplot.maxstat$maximal.respiration[[2]], TTTplot.maxstat$maximal.respiration[[3]],
          TTTplot.maxstat$spare.respiratory.capacity[[2]], TTTplot.maxstat$spare.respiratory.capacity[[3]],
          OSplot.maxstat$glycolytic.capacity[[2]], OSplot.maxstat$glycolytic.capacity[[3]],
          OSplot.maxstat$glycolytic.reserve[[2]], OSplot.maxstat$glycolytic.reserve[[3]],
          labels = "AUTO", ncol = 2)
```


#### Using median

Time to next treatment (median).
```{r}
TTTplot.med <- list()

for (seaName in rownames(seaCombat)) {
  seaName.format <- formatSea(seaName)
  pvals <- c(ttt[seaName.format ,"p"], ttt.M[seaName.format ,"p"], ttt.U[seaName.format ,"p"])
  pvals <- sapply(pvals, function(x) {
    sprintf("p value = %1.2f",x)
  })
  TTTplot.med[[seaName]] <- km(survT, seaName = seaName, split = "IGHV", t="TTT",
   titlePlot =c("(IGHV", "umutated)", "mutated)"),  hr="tr", c="med", pvals = pvals)
}
```

Overall survival (median).
```{r}
OSplot.med <- list()

for (seaName in rownames(seaCombat)) {
  seaName.format <- formatSea(seaName)
  pvals <- c(os[seaName.format ,"p"], os.M[seaName.format ,"p"], os.U[seaName.format ,"p"])
  pvals <- sapply(pvals, function(x) {
    sprintf("p value = %1.2f",x)
  })
  OSplot.med[[seaName]] <- km(survT, seaName = seaName, split = "IGHV", t="OS",
   titlePlot =c("(IGHV", "umutated)", "mutated)"),  hr="tr", c="med", pvals = pvals)
}
```

KM plot for all samples (Figure 6)
```{r TTT_OS_combine_median, fig.path=plotDir, fig.width = 12, fig.height = 8, dev = c("png", "pdf")}
title = ggdraw() + draw_figure_label("Figure 6", fontface = "bold", position = "top.left",size=20)
p<-plot_grid(TTTplot.med$glycolytic.reserve[[1]], TTTplot.med$maximal.respiration[[1]],
          TTTplot.med$spare.respiratory.capacity[[1]],
          OSplot.med$glycolytic.capacity[[1]], OSplot.med$glycolytic.reserve[[1]],
          labels = NULL)
plot_grid(title, p, rel_heights = c(0.05,0.95), ncol = 1)
```

KM plot for IGHV stratified samples (not shown in manuscript)
```{r TTT_OS_IGHV_UM_median, fig.path=plotDir, fig.width = 10, fig.height = 20, dev = c("png", "pdf")}
plot_grid(TTTplot.med$glycolytic.reserve[[2]], TTTplot.med$glycolytic.reserve[[3]],
          TTTplot.med$maximal.respiration[[2]], TTTplot.med$maximal.respiration[[3]],
          TTTplot.med$spare.respiratory.capacity[[2]], TTTplot.med$spare.respiratory.capacity[[3]],
          OSplot.med$glycolytic.capacity[[2]], OSplot.med$glycolytic.capacity[[3]],
          OSplot.med$glycolytic.reserve[[2]], OSplot.med$glycolytic.reserve[[3]],
          labels = "AUTO", ncol = 2)
```


## Multi-variate cox model

### Prepare dataset
```{r}
seaTable <- assays(seaCombat)$seaMedian

survT = patmeta[colnames(seaTable),]
survT[which(survT[,"IGHV"]=="U") ,"IGHV"] = 0
survT[which(survT[,"IGHV"]=="M") ,"IGHV"] = 1
survT$IGHV = as.numeric(survT$IGHV)

colnames(survT) = gsub("Age4Main", "age", colnames(survT))

#add seahorse measurement information

survT <- cbind(survT, t(mscale(seaTable)))

# competinting risk endpoint fpr 
survT$compE <- ifelse(survT$treatedAfter == TRUE, 1, 0)
survT$compE <- ifelse(survT$treatedAfter == FALSE & survT$died==TRUE,
                      2, survT$compE )
survT$T7  <- ifelse(survT$compE == 1, survT$T5, survT$T6 )

#genetic variants
survT$SF3B1      <- Biobase::exprs(lpdAll)[ "SF3B1",      rownames(survT)  ]
survT$NOTCH1     <- Biobase::exprs(lpdAll)[ "NOTCH1",     rownames(survT)  ]
survT$BRAF       <- Biobase::exprs(lpdAll)[ "BRAF",       rownames(survT)  ]
survT$TP53       <- Biobase::exprs(lpdAll)[ "TP53",       rownames(survT)  ]
survT$del17p13   <- Biobase::exprs(lpdAll)[ "del17p13",   rownames(survT)  ]
survT$del11q22.3 <- Biobase::exprs(lpdAll)[ "del11q22.3", rownames(survT)  ]
survT$trisomy12 <-  Biobase::exprs(lpdAll)[ "trisomy12", rownames(survT)  ]
```

```{r extract}
extractSome <- function(x) {
  sumsu <- summary(x)
  data.frame(
    `p-value`      = 
      sprintf("%6.3g", sumsu[["coefficients"]][, "Pr(>|z|)"]),
    `HR`           = 
      sprintf("%6.3g", signif( sumsu[["coefficients"]][, "exp(coef)"], 2) ), 
    `lower 95% CI` = 
      sprintf("%6.3g", signif( sumsu[["conf.int"]][, "lower .95"], 2) ),
    `upper 95% CI` = 
      sprintf("%6.3g", signif( sumsu[["conf.int"]][, "upper .95"], 2),
              check.names = FALSE) )
}
```

Define covariates and effects.
```{r covariates, echo=FALSE}
survT$age <- survT$age/10
survT$IC50beforeTreatment <- ifelse(survT$IC50beforeTreatment==TRUE, 1, 0)
survT$IGHVwt <- ifelse(survT$IGHV==1, 0, 1)
```

#### TTT

glycolytic reserve 
```{r}
surv1 <- coxph(
  Surv(T5, treatedAfter) ~  
    age +
    as.factor(IC50beforeTreatment) +
    as.factor(trisomy12) +
    as.factor(del11q22.3) +
    as.factor(del17p13) +
    as.factor(TP53) +
    IGHVwt +
    glycolytic.reserve,       # continuous
  data = survT )

colFactor <- data.frame(factor = c("age", "pretreatment", 
                                   "trisomy12", "del11q22.3", 
                                   "del17p13","TP53","U-CLL",
                                   "glycolytic reserve"))

outTab <- cbind(colFactor,extractSome(surv1))

write(print(xtable(outTab,
            caption = "Multivariate Cox regression model for time to treatment with glycolytic reserve as a covariate"), 
      include.rownames=FALSE,
      caption.placement = "top"), file = "section05/glyRes_TTT.tex")


cat(sprintf("%s patients considerd in the model; number of events %1g\n", 
            summary(surv1)$n, summary(surv1)[6] ) )
```

maximal respiration 
```{r}
surv1 <- coxph(
  Surv(T5, treatedAfter) ~  
    age +
    as.factor(IC50beforeTreatment) +
    as.factor(trisomy12) +
    as.factor(del11q22.3) +
    as.factor(del17p13) +
    as.factor(TP53) +
    IGHVwt +
    maximal.respiration,       # continuous
  data = survT )

colFactor <- data.frame(factor = c("age", "pretreatment", 
                                   "trisomy12", "del11q22.3", 
                                   "del17p13","TP53","U-CLL",
                                   "maximal respiration"))

outTab <- cbind(colFactor,extractSome(surv1))

write(print(xtable(outTab,
            caption = "Multivariate Cox regression model for time to treatment with maximal respiration as a covariate"), 
      include.rownames=FALSE,
      caption.placement = "top"), file = "section05/maxRes_TTT.tex")


cat(sprintf("%s patients considerd in the model; number of events %1g\n", 
            summary(surv1)$n, summary(surv1)[6] ) )
```

spare respiratory capacity
```{r}
surv1 <- coxph(
  Surv(T5, treatedAfter) ~  
    age +
    as.factor(IC50beforeTreatment) +
    as.factor(trisomy12) +
    as.factor(del11q22.3) +
    as.factor(del17p13) +
    as.factor(TP53) +
    IGHVwt +
    spare.respiratory.capacity,       # continuous
  data = survT )

colFactor <- data.frame(factor = c("age", "pretreatment", 
                                   "trisomy12", "del11q22.3", 
                                   "del17p13","TP53","U-CLL",
                                   "spare.respiratory.capacity"))

outTab <- cbind(colFactor,extractSome(surv1))

write(print(xtable(outTab,
            caption = "Multivariate Cox regression model for time to treatment with glycolytic reserve as a covariate"), 
      include.rownames=FALSE,
      caption.placement = "top"), file = "section05/spResCap_TTT.tex")


cat(sprintf("%s patients considerd in the model; number of events %1g\n", 
            summary(surv1)$n, summary(surv1)[6] ) )
```

### OS

glycolytic reserve 
```{r}
surv1 <- coxph(
  Surv(T6, died) ~  
    age +
    as.factor(IC50beforeTreatment) +
    as.factor(trisomy12) +
    as.factor(del11q22.3) +
    as.factor(del17p13) +
    as.factor(TP53) +
    IGHVwt +
    glycolytic.reserve,       # continuous
  data = survT )

colFactor <- data.frame(factor = c("age", "pretreatment", 
                                   "trisomy12", "del11q22.3", 
                                   "del17p13","TP53","U-CLL",
                                   "glycolytic reserve"))

outTab <- cbind(colFactor,extractSome(surv1))

write(print(xtable(outTab,
            caption = "Multivariate Cox regression model for overall survival with glycolytic reserve as a covariate"), 
      include.rownames=FALSE,
      caption.placement = "top"), file = "section05/glyRes_OS.tex")


cat(sprintf("%s patients considerd in the model; number of events %1g\n", 
            summary(surv1)$n, summary(surv1)[6] ) )
```

glycolytic capacity 
```{r}
surv1 <- coxph(
  Surv(T6, died) ~  
    age +
    as.factor(IC50beforeTreatment) +
    as.factor(trisomy12) +
    as.factor(del11q22.3) +
    as.factor(del17p13) +
    as.factor(TP53) +
    IGHVwt +
    glycolytic.capacity,       # continuous
  data = survT )

colFactor <- data.frame(factor = c("age", "pretreatment", 
                                   "trisomy12", "del11q22.3", 
                                   "del17p13","TP53","U-CLL",
                                   "glycolytic capacity"))

outTab <- cbind(colFactor,extractSome(surv1))

write(print(xtable(outTab,
            caption = "Multivariate Cox regression model for overall survival with glycolytic capacity as a covariate"), 
      include.rownames=FALSE,
      caption.placement = "top"), file = "section05/glyCap_OS.tex")


cat(sprintf("%s patients considerd in the model; number of events %1g\n", 
            summary(surv1)$n, summary(surv1)[6] ) )
```
