---
title: 'Section 1: Comparison between normal B cell and CLL cells'
author: "Junyan Lu"
date: "`r doc_date()`"
output:
  BiocStyle::html_document
---

# Comparison between normal B cell and CLL cells

```{r, message=FALSE, include=!exists(".standalone"), eval=!exists(".standalone"),echo=FALSE}
library(seahorseCLL)
library(BloodCancerMultiOmics2017)
library(SummarizedExperiment)
library(ggbeeswarm)
library(xtable)
library(cowplot)
library(gridExtra)
library(tidyverse)
```

```{r echo=FALSE}
plotDir = ifelse(exists(".standalone"), "", "section01/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
options(stringsAsFactors=FALSE)
```


## A global view of bioenergetic features in CLL and normal B-cells

Load data
```{r}
data("seaBcell", "seaOri", "patmeta")
```

Combine the two data sets to one matrix
```{r}
stopifnot(rownames(seaBcell) == rownames(seaOri))
seaOri$diagnosis <- patmeta[colnames(seaOri),]$Diagnosis
seaOri <- seaOri[,seaOri$diagnosis == "CLL"] # choose CLL samples

seaMat <- cbind(assays(seaBcell)$seaMedian, assays(seaOri)$seaMedian)
seaBatch <- rbind(colData(seaBcell)[,"dateMST", drop= FALSE], colData(seaOri)[,"dateMST", drop = FALSE])

#remove samples that contain NA values
seaMat <- seaMat[,complete.cases(t(seaMat))]
```

Principal component analysis
```{r}
resPC <- prcomp(t(seaMat), center = TRUE, scale. = TRUE)
varExp <- resPC$sdev^2/sum(resPC$sdev^2)
```

Plot the first two principal components
```{r, fig.width=5, fig.height=5}
#define color
colorList <- c(`B cell` = "#FF3030", CLL= "#1E90FF")

plotTab <- data.frame(resPC$x[,c(1,2)])
plotTab$type <- c(rep("B cell", ncol(seaBcell)), rep("CLL", nrow(plotTab)- ncol(seaBcell))) #10 normal b cell samples

pcaPlot <- ggplot(plotTab, aes(x=PC1, y=PC2, color = type)) + geom_point(size=3) + 
  xlab(sprintf("PC1 (%2.1f%s)",varExp[1]*100,"%")) + ylab(sprintf("PC2 (%2.1f%s)", varExp[2]*100, "%")) + 
  theme_bw() + theme(legend.position = c(0.9,0.9), legend.title = element_blank(), 
                     legend.background = element_rect(color="grey"),
                     axis.text = element_text(size =18),
                     axis.title = element_text(size = 20)) +
  scale_color_manual(values = colorList) + coord_cartesian(xlim = c(-5.5,5.5), ylim = c(-5.5,5.5))
pcaPlot
```

## Associations between cell type and individual metabolic features

Prepare table for hypothesis test
```{r}
seaMat <- data.frame(cbind(assays(seaBcell)$seaMedian, assays(seaOri)$seaMedian)) %>% rownames_to_column(var = "measure")
seaTab <- gather(seaMat, key = "patientID", value = "value", -measure) %>% 
  mutate(type = ifelse(substr(patientID, 1, 1) == "K", "B cell", "CLL")) %>% mutate(type = factor(type)) %>%
  mutate(batch = as.factor(seaBatch[patientID, ]))
```

t-test for each measurment
```{r}
pTab <- group_by(seaTab, measure) %>% do((function(x) {
  res <- t.test(value ~ type, x, equal.var = TRUE)
  data.frame(p = res$p.value,
             diff = res$estimate[[2]] - res$estimate[[1]])
}) (.))
pTab$p.adj <- p.adjust(pTab$p, method = "BH")
```

ANOVA-test for each measurment (accounting for batch effect)
```{r}
pTab.aov <- group_by(seaTab, measure) %>% do((function(x) {
  res <- summary(aov(value ~ type + batch, x))
  data.frame(p = res[[1]][["Pr(>F)"]][1])
}) (.))
pTab.aov$p.adj <- p.adjust(pTab.aov$p, method = "BH")
```

Expor the table to LaTex format using xtable
```{r, results=FALSE}
expTab <- pTab %>% ungroup() %>% mutate(measure = gsub("\\."," ",measure)) %>%
  rename("Seahorse measurement" = measure, "Difference of mean" = diff) %>%
  mutate(p = pTab.aov$p,  `adjusted p` = pTab.aov$p.adj) %>%
  select(-p.adj)
expTab[expTab$`Seahorse measurement` == "ECAR OCR ratio", 1] <- "ECAR/OCR"
fileConn <- file("section01/tTest_BcellVSCLL.tex")
writeLines(print(xtable(expTab, digits = 3, 
             caption = "ANOVA test results of energy metabolic measurements between CLL cells and normal B cells "), 
      include.rownames=FALSE,
      caption.placement = "top"), fileConn)
close(fileConn)
```

Beeswarms plot for select measurement
```{r, warning=FALSE}
measureList <- c("basal.respiration","glycolysis","ATP.production","glycolytic.capacity","maximal.respiration","glycolytic.reserve")
gList <- lapply(measureList, function(seaName) {
  
  plotTab <- filter(seaTab, measure == seaName, !is.na(value))
  pval <- filter(pTab.aov, measure == seaName )$p
  
    
  #unit y (add unit to y axis, based on the type of measurement)
  if (seaName %in% c("basal.respiration","ATP.production","maximal.respiration")) {
    yLab <- "OCR (pMol/min)"
  } else yLab <- "ECAR (pMol/min)"
    
  #replace the "." in the mearement name with space
  seaName <- gsub("\\."," ",seaName)

  
  #plot title
  #plotTitle <- sprintf("p value = %s", format(pval, digits = 2, scientific = TRUE))
    plotTitle <- paste(sprintf("'%s (p = '~",seaName),
                     sciPretty(pval, digits = 2),"*')'")
  ggplot(plotTab, aes(x=type, y = value)) + 
       stat_boxplot(geom = "errorbar", width = 0.3) +
       geom_boxplot(outlier.shape = NA, col="black", width=0.4) + 
       geom_beeswarm(cex=2, size =0.5, aes(col = type)) + theme_classic() +
       xlab("") + ylab(yLab) + ggtitle(parse(text = plotTitle)) +
      theme(axis.line.x = element_blank(), axis.ticks.x = element_blank(),
             axis.title.y = element_text(size=10, face="bold"),
             axis.text = element_text(size=11),
             plot.title = element_text(size = 12, hjust=0.5),
             legend.position = "none",
             axis.text.x = element_text(face="bold",size=12)) +
    scale_color_manual(values = colorList)
  
})
beePlot <- grid.arrange(grobs = gList, ncol=2)
```


Combine the PCA plot and beeswarm plots (Figure 1)
```{r normalVScll, fig.path=plotDir, dev=c("png", "pdf"), fig.width=12, fig.height=6}
plot_grid(pcaPlot, beePlot, labels= c("A","B"), rel_widths = c(1,1))
```


