---
title: 'Section 4: Transcriptomic analysis'
author: "Junyan Lu"
date: "`r doc_date()`"
output:
  BiocStyle::html_document
---
# Transcriptomic analysis
```{r, message=FALSE, include=!exists(".standalone"), eval=!exists(".standalone"), echo=FALSE}
library(seahorseCLL)
library(BloodCancerMultiOmics2017)
library(DESeq2)
library(genefilter)
library(ggbeeswarm)
library(SummarizedExperiment)
library(RColorBrewer)
library(piano)
library(pheatmap)
library(grid)
library(gridExtra)
library(cowplot)
library(tidyverse)
```

```{r echo=FALSE}
plotDir = ifelse(exists(".standalone"), "", "section04/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
options(stringsAsFactors=FALSE)
```


## Differential expression between M-CLL and U-CLL samples

### Data pre-processing
Load data set
```{r}
data("dds", "patmeta","mutCOM","seaOri")
```

Only use CLL samples with IGHV annotations and have been used in seahorse experiments
```{r}
#annotate IGHV status
dds$IGHV <- patmeta[match(dds$PatID, rownames(patmeta)),]$IGHV
dds$diag <- patmeta[match(dds$PatID, rownames(patmeta)),]$Diagnosis

#estimate size factor
dds <- estimateSizeFactors(dds)

#only choose CLL samples with IGHV annotations and have been used in seahorse experiments
ddsCLL <- dds[,dds$diag == "CLL" & !is.na(dds$IGHV) & dds$PatID %in% colnames(seaOri)]

#how many samples do we have?
ncol(ddsCLL)
```

Filter genes
```{r}
#remove genes without gene symbol annotations
ddsCLL <- ddsCLL[! rowData(ddsCLL)$symbol %in% c(NA,""),]
ddsCLL <- ddsCLL[rowData(ddsCLL)$chromosome != "Y",]

#only keep genes that have counts higher than 10 in any sample
keep <- apply(counts(ddsCLL), 1, function(x) any(x >= 10)) 
ddsCLL <- ddsCLL[keep,]

# Remove transcripts which do not show variance across samples.
sds <- rowSds(counts(ddsCLL, normalized = TRUE))
sh <- shorth(sds)
ddsCLL <- ddsCLL[sds >= sh,]

#how many genes do we have
nrow(ddsCLL)
```

Variance stabilizing tranformation
```{r, cache=TRUE}
ddsCLL.norm <- varianceStabilizingTransformation(ddsCLL)
```

### Identify differentially expressed gene between M-CLL and U-CLL using DESeq2

Differential expression using DESeq2
```{r, cache=TRUE}
ddsCLL$IGHV <- factor(ddsCLL$IGHV, levels = c("U", "M"))
design(ddsCLL) <- ~ IGHV
ddsCLL <- DESeq(ddsCLL, betaPrior = TRUE)
```

Get differential expression result
```{r}
DEres <- as.tibble(results(ddsCLL, tidy = TRUE)) %>% mutate(symbol = rowData(ddsCLL)$symbol)
```

## Gene enrichment analysis

Function for converting DEseq results to enrichment analysis input
```{r}
createInput <- function(DEres, pCut = 0.05, ifFDR = FALSE, rankBy = "stat") {
  if (ifFDR) {
    inputTab <- filter(DEres, padj <= pCut)
  } else {
    inputTab <- filter(DEres, pvalue <= pCut)
  }
  
  inputTab <- arrange(inputTab, pvalue) %>% filter(!duplicated(symbol)) %>% select_("symbol", rankBy) %>% data.frame(stringsAsFactors = FALSE)
  rownames(inputTab) <- inputTab$symbol
  inputTab$symbol <- NULL
  colnames(inputTab) <- "stat"
  return(inputTab)
}
```

load genesets
```{r}
gmts = list(H=system.file("extdata","h.all.v5.1.symbols.gmt", package="BloodCancerMultiOmics2017"),
            C6=system.file("extdata","c6.all.v5.1.symbols.gmt", package="BloodCancerMultiOmics2017"),
            KEGG=system.file("extdata","c2.cp.kegg.v5.1.symbols.gmt", package = "BloodCancerMultiOmics2017"))
```

Enrichment analysis using Hallmarks gene set
```{r, warning=FALSE, message=FALSE}
enRes <- list()
inputTab <- createInput(DEres, pCut = 0.1, ifFDR = TRUE)
enRes[["Gene enrichment analysis"]] <- runGSEA(inputTab = inputTab, gmtFile = gmts$H, GSAmethod = "page")

#remove the HALLMARK_
enRes$`Gene enrichment analysis`$Name <- gsub("HALLMARK_","", enRes$`Gene enrichment analysis`$Name)
```

Plot hallmark result
```{r, fig.width=8, fig.height=4}
enBar <- plotEnrichmentBar(enRes, pCut = 0.1, ifFDR = TRUE, setName = "Hallmark gene sets")
plot(enBar)
```

### Heatmap plot of the expression values of genes in the glycolysis geneset

Prepare the data for heatmap
```{r}
# load genes in the gene set
gsc <- loadGSC(gmts$H)
geneList <- gsc$gsc$HALLMARK_GLYCOLYSIS

#select differentially expressed genes
fdrCut <- 0.10
sigDE <- filter(DEres, padj <= fdrCut, log2FoldChange < 0) %>% filter(symbol %in% geneList) %>%
  arrange(log2FoldChange)

#get the expression matrix
plotMat <- assay(ddsCLL.norm[sigDE$row,])
colnames(plotMat) <- ddsCLL.norm$PatID
rownames(plotMat) <- sigDE$symbol

#sort columns of plot matrix based on trisomy12 status
plotMat <- plotMat[,order(ddsCLL$IGHV)]

#calculate z-score and sensor
plotMat <- t(scale(t(plotMat)))
plotMat[plotMat >= 4] <- 4
plotMat[plotMat <= -4] <- -4

annoCol <- data.frame(row.names = ddsCLL.norm$PatID, `IGHV` = ddsCLL.norm$IGHV)
```

Plot the heatmap
```{r, fig.height=8, fig.width=8}
#color for colum annotation
annoColor <- list(IGHV = c(M = "red", U = "grey80"))

hallHeatmap <- pheatmap(plotMat, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
         cluster_cols = FALSE, cluster_rows = FALSE,
         annotation_col = annoCol, annotation_colors = annoColor,
         show_colnames = FALSE, fontsize_row = 8, breaks = seq(-5,5, length.out = 101), treeheight_row = 0,
         border_color = NA, main = "HALLMARK_GLYCOLYSIS",silent = TRUE)$gtable

grid.draw(hallHeatmap)
```

### Beeswarm plot of expression values of key glycolytic genes

```{r}
plotGenes <- c("PFKP","PGAM1","PGK1")
plotTab <- data.frame(assay(ddsCLL.norm[rowData(ddsCLL.norm)$symbol %in% plotGenes,]))
colnames(plotTab) <- ddsCLL.norm$PatID
plotTab$symbol <- rowData(ddsCLL.norm[rowData(ddsCLL.norm)$symbol %in% plotGenes,])$symbol
plotTab <- gather(plotTab, key = "PatID", value = "value", -symbol) %>%
  mutate(IGHV = colData(ddsCLL.norm)[match(PatID, ddsCLL.norm$PatID),]$IGHV)

beePlot <- ggplot(plotTab, aes(x=IGHV, y = value)) + 
    stat_boxplot(geom = "errorbar", width = 0.3) +
    geom_boxplot(outlier.shape = NA, col="black", width=0.4) + 
    geom_beeswarm(cex=2, size =0.5, aes(col = IGHV)) + theme_classic() +
    xlab("") + ylab("normalized expression") +
    scale_color_manual(values = c("M" = "red","U" = "grey30")) +
    theme(axis.line.x = element_blank(), axis.ticks.x = element_blank(),
           axis.title = element_text(size=12, face="bold"),
           axis.text = element_text(size=12),
           plot.title = element_text(face="bold", hjust=0.5),
           legend.position = "none",
           axis.title.x = element_text(face="bold"),
          strip.background = element_blank(),
          strip.text = element_text(size=15, face = "bold")) +
  facet_wrap(~ symbol, scales = "free")
```

### Combined plots for manuscript (Figure 3)
```{r expressionIGHV, fig.height=6, fig.width=10, fig.path=plotDir, dev=c("png", "pdf")}
ggdraw() + 
  draw_plot(enBar, 0, 0.5, 0.5, 0.45) + 
  draw_plot(hallHeatmap, 0.5, 0, 0.5, 0.95) +
  draw_plot(beePlot, 0, 0, 0.5, 0.4) +
  draw_plot_label(c("A","B","C"), c(0, 0.5, 0), c(1, 1, 0.45), fontface = "plain", size=20)
```
